# Reproducible Research - Project 1
#### Chris Holman

### Question 1
It is important to ensure the working directory is correct and read in the file. In this case, we used the `read_csv()` which is party of the tidyverse package. Notice we loaded the library before reading in our data. 
```{r Reading in dataset}
getwd()

library(tidyverse)

activity <- read_csv("./activity.csv")

activity_original <- activity

head(activity)
str(activity)


```
The variables included in this dataset are:

-steps: Number of steps taking in a 5-minute interval (missing values are coded as **NA**)
-date: The date on which the measurement was taken in YYYY-MM-DD format
-interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.
### Question 2 
Ignore any missing values 

```{r}
cleanActivity <- activity[!(is.na(activity$steps)),]
``` 
Aggregte steps by date
```{r}
dailyActivity <- aggregate(steps ~ date, cleanActivity, sum)
```

Creating a histogram for number of steps taken each day.
```{r Histogram of total numer of steps taken each day}
hist(dailyActivity$steps, breaks = 20, xlab = "Steps Taken",
     main = "Total Number of Steps Taken per Day")
```


```{r Mean and Median of number of steps taken each day}
library(xtable)
library(dplyr)
stepMean <- mean(dailyActivity$steps)
stepMedian <- median(dailyActivity$steps)
summary <- summary(dailyActivity)
```
The mean calculated is *`stepMean`* and the median is *`stepMedian`*. 
### Question 3

To calculate the number of steps taken per interval, we need to aggregate by interval

```{r}
intervalActivity <- aggregate(steps ~ interval, cleanActivity, mean)

```

```{r Time series plot of the average number of stps taken}

stepPlot <- plot(x = intervalActivity$interval, y = intervalActivity$steps, type = "l",
		 main = "Time Series Plot of Steps taken per Interval",
		 ylab = "Number of Steps",
		 xlab = "Intervals (5 mins)"
		 )  
	
```
Which five minute interval contains the maximum number of steps?
```{r 5-minute interval that, on average, contains teh maximum number of steps}
maxSteps <- intervalActivity[grep(max(intervalActivity$steps), intervalActivity$steps),]
```
The interval with the most steps is `maxSteps$interval`, with `maxSteps$steps` steps.

### Question 4

Replace NA values in Steps with mean value for that interval


```{r Describe and show a strategy for imputing missing data}
imputedActivity <- activity
for(x in 1:17568){
	if(is.na(imputedActivity[x, 1])==TRUE){
		imputedActivity[x,1] <- intervalActivity[intervalActivity$interval %in% imputedActivity[x,3],2]
	}
}
```

Aggregate steps by day in the new data with imputed step values

```{r}
imputedActivityDailySteps <- aggregate(steps ~ date, imputedActivity, sum)
```


Create histogram of new imputed data.

```{r Histogram of the toal number of steps taeken each day after missing values are imputed}
hist(imputedActivityDailySteps$steps, breaks = 20, 
     xlab = "Steps Taken",
     main = "Histogram of Total Number of Steps Taken per day (Imputed Values)")
```

Calculate the mean an median total number of steps each day. 

```{r}
imputedStepsSummary <- summarise(imputedActivityDailySteps,
				 meanImputedSteps = mean(imputedActivityDailySteps$steps),
				 medianImputedSteps = median(imputedActivityDailySteps$steps))
```

The mean of the data including imputed steps was `meanImputedSteps` steps and the median was `medianImputedSteps` steps. 

## Question 5

To compare the average number of steps we must format a new variable to categorize the observation as part of the weekday or weekend.


```{r Panel plot comparing the average number of steps taken per 5- minute interval across week days and weekends}
activityDays <- imputedActivity
activityDays$days <- weekdays(activityDays$date)
activityDays$weekday <- as.character(rep(0, times=17568))
for(x in 1:17568) {
    if(activityDays[x, 4] %in% c("Saturday", "Sunday")) {
        activityDays[x, 5] <- "weekend"
    } else {
        activityDays[x, 5] <- "weekday"
    }
}
activityDays$weekday <- factor(activityDays$weekday)

```

We can will separate the data into a weekday and weekend set.

```{r}
weekdayData <- activityDays[activityDays$weekday == "weekday",]
weekendData <- activityDays[activityDays$weekday == "weekend",]
```

Then group the data by interval by the average number of steps

```{r}
weekdayMean <- aggregate(steps ~ interval, weekdayData, mean)
weekendMean <- aggregate(steps ~ interval, weekendData, mean)
```

Last, create the final plot comparing the average number of steps taken per 5 min interval across weekdays and weekend. 

```{r Panel plot comparing the average number of steps taken per 5- minute interval across weekdays and weekends}

par(mfrow=c(2, 1), mar=c(4, 4.1, 3, 2.1))
plot(weekdayMean$interval, weekdayMean$steps, type="l",
     main="Time Series Plot of Average Steps Taken per Interval, for Weekdays",
     xlab="Intervals (in 5 mins)", ylab="Number of Steps", col = "red")
plot(weekendMean$interval, weekendMean$steps, type="l",
     main="Time Series Plot of Average Steps Taken per Interval, for Weekends",
     xlab="Intervals (in 5 mins)", ylab="Number of Steps", col = "green")


```






















